---
title: "Reading and writing data"
output: html_document
---


This is partly based an excellent tutorial [R for SAS and SPSS users](https://science.nature.nps.gov/im/datamgmt/statistics/r/documents/r_for_sas_spss_users.pdf)


Let's start with a simple example...

We have a dataset in a text file. The variables are separated with tabs (empty space). The names of the variables names are on the first row. R will try to guess whether a text file has variable names in the header or not, but quite often it fails. Adding header=TRUE 
to the function call (see below) helps:

```{r}

apgar <- read.table("apgar.txt", header=TRUE)
head(apgar) # we used to preview only the start or head of the data frame

```
That was not complicated! However, there are a couple things you need to pay attention to. 

*Working directory*

The filename is resolved relative to the current working directory of your R console. If the file is in the working directory it can be found with the filename. If not, you either need to copy the file to your working directory or give also the path to the file. 

But what is your working directory? You get to know it with the *getwd* command. 

```{r}
getwd()

```

When you first start Rstudio, it is good idea to change the working directory to the directory where the scripts and data of your current project are located. The easiest way 
is to use the Files-browser in Rstudio, go to the right directory, and the click the cog 
(hammasratas) called "More" and choose >> Set As Working Directory.

You can also set the working directory manually via the command line:

```{r}
setwd("~/github/kickstartr")  # write the right path 

```

*Separator used*

By default, read.table uses any whitespace character to separate different variables on rows. Often the separator is something else, like a comma. In apgar.csv the same data is in CSV format (CSV = comma separated values), and the simple call to read.table fails. Try this and have a look at the data: it's extremely useful for you to know what the data frame looks
like when the wrong separator is used. Sooner or later you will encounter this kind of a problem in your data analysis. 

```{r}
# apgar <- read.table("apgar.csv")
# head(apgar)
```

The problem is easily corrected by specifying the separator correctly:
```{r}
apgar <- read.table("apgar.csv", sep=",", header=TRUE)
head(apgar)
```

Alternatively, you can use function read.delim, read.csv etc. They are just shortcuts to read.table with slightly different default values for the parameters (such as the separator
character). 

*Decimals are marked with a comma, not with a dot*

In Finnish, the convention is to use decimal comma, not the decimal dot. By default SPSS, Excel etc. detect the locale settings of the operating system and adapt their notation accordingy. This creates sometimes problems with other data analysis software, which often follow the English convention and the decimal comma. 

```{r}
# Separator is here ;
apgar <- read.table("apgar_decimalcomma.txt", sep=";", header=TRUE)
head(apgar)
```

At first, everything seems to be fine - at least until you try to use some of the decimal values. 

```{r}
apgar$lgincome * 2
```

This can solved by specifying the decimal point character:
```{r}
apgar <- read.table("apgar_decimalcomma.txt", dec=",", sep=";", header=TRUE)
head(apgar)
```


Reading SPSS files
------------------------
R can read for example SPSS-files directly, using the package *foreign*

```{r}
library(foreign)
apgar <- read.spss("apgar.sav",to.data.frame=TRUE)

View(apgar)
```


Writing data
---------------------------
All the above presented text data files were actually created from the aforementioned apgar.sav file. The following functions calls were used. 

```{r}
# first load the data
library(foreign) 
apgar <- read.spss("apgar.sav",to.data.frame=TRUE)

# then save it
write.table(apgar, "apgar.txt", sep="\t", row.names=FALSE) # use tabs
write.table(apgar, "apgar.csv", sep=",", row.names=FALSE) # use commas
write.table(apgar, "apgar_decimalcomma.txt", sep=";", dec=',', row.names=FALSE) 

# note the col.names=FALSE
# by default, R writes also row.names
# Typically row.names are equal to a consequtive numbering of the lines, 
# which is not very useful to save.

```

