---
title: "Representing data and performing simple calculations"
output: html_document
---


Objects are a way to represent data in a programming language like R. For different kinds of data, the computer needs different kind of objects. 

Vectors
==========================

In R, the basic object is vector. It is an ordered set of data containing one or more elements. 

When you enter text to the R console, the interpreter will parse your input, and create a vector. The vectors "know" what kind of data they have in them.

[R lang definitions: Vectors](https://cran.r-project.org/doc/manuals/r-release/R-lang.html#Vector-objects)


```{r}

# creates a vector, which contains only one number
3    

# creates a vector with three members
c(7, 6, 3)

# the vectors have type
class(3)
class(c(7, 6, 3))

# and length
length(3)
length(c(7, 6, 3))


# creates object of type character
"Vector"
c("Vector", "says", "hello!")

# beside the type, everything is the same as with numbers
class("Vector")
class(c("Vector", "says", "hello!"))

length("Vector")
length(c("Vector", "says", "hello!"))

```

Variables 
=========================

The power of using programming languages stems from the possibility of describing the *algorithm* (= what will be computed) regardless of what the actual data looks like. The actual data, on its part, is stored in *variables*, which represent the actual objects in the algorithm. 

```{r}

salaryBeforeTaxes <- 2734.51
employerCosts <- salaryBeforeTaxes * 1.56 
print(employerCosts)

```

Here the R console first creates a vector of type *numerical*. The vector is of length one, it has only a single member (2734.51). The vector is then assigned to be the value of the variable salaryBeforeTaxes using the arrow operator "<-". You can think of variables as boxes in which you can put data. 


Operators
===========================

Data is contained in objects; variables represent the data in algorithms. In the previous example, we performed a *multiplication* operiation which was represented using \* as an operator. When an operator encounters objects (or variables which can represent them), new objects are created. 

Even though the idea is a bit abstract, the simplest examples of it are familiar already from elementary school mathematics.

[R lang definition: Operators](https://cran.r-project.org/doc/manuals/r-release/R-lang.html#Operators)


```{r}
# Let's create objects and assign them to variables.
a <- 4
b <- 6
c <- -3

a + b
a - b
a * b
a / b

# NB. Order of calculation follows the ordinary mathematical rules
a + b * c    
(a + b) * c  
a + (b * c) 

```

These were vectors of length 1. Everything happens pretty much the same way for vectors which are longer. In that case, the operator is applied element-wise. 

```{r}
thesisWorker <- 0
researchAssistant <- 1700
doctoralStudent <- 2200
postdoc <- 3200
professor <- 5500
rector <- 25000

salaryBeforeTaxes <- c(thesisWorker, 
                       researchAssistant,
                       doctoralStudent,
                       postdoc,
                       professor,
                       rector)

employerCosts <- salaryBeforeTaxes * 1.56 
employerCosts

```

Variables are a way to describe algorithms irrespective of data. Vectors are a very powerful way to calculate the same thing for multiple values at once. 


Data frames
==============================

Most of the time researchers in behavioural sciences are working with data which are nicely arranged as two dimensional datasets. On the rows, there are observations and on columns there are different variables which describe different aspects of these observations. In R, these are called "data frame objects" or just "data frames". 

Data frames can be constructed by stacking vectors beside each other or on top of each other


```{r}

position <- c("Thesis worker", 
               "Research assistant",
               "Doctoral student",
               "Postdoc", 
               "Professos", 
               "Rector")
salaryBeforeTaxes <- c(thesisWorker, 
                       researchAssistant,
                       doctoralStudent,
                       postdoc,
                       professor,
                       rector)
workingYears <- c(0, 1, 3, 8, 20, 30)
citations <- c(0, 
               1,
               8,
               30,
               7784,
               2021)
academics <- data.frame(position, salaryBeforeTaxes, citations, workingYears)

```

We can view this nice data frame either printing it to the console or using View. 

```{r}
academics

View(academics)

```



Dots . in the variable or function names
----------------------------------------

In the previous example, we saved the data frame with a variable name which had an ending ".df". It is good to notice that there is nothing magical here. It is just a part of the variable name, and we could have used any other valid variables name. Unlike many other programming languages such as Python, Java or MATLAB, dot (.) is not used for accessing object/struct members. R uses $-symbol for that purpose. 



Accessing the variables in data frames
-----------------------------------------

Vectors (or variables) within data frames can be easily accessed and used with $.  

```{r}

academics$salaryBeforeTaxes
```

The variables in the data frames behave like vectors with operators.  

```{r}
academics$salaryBeforeTaxes * 1.56
```


Using function "names", it is also possible to get all the variable names as a vector. 

```{r}
# get the names as a vector
names(academics)  

```

This can be also used to change their names.
```{r}
# set the names from a vector
academics.cryptic <- academics # make a copy of the data frame first

names(academics.cryptic)

# Use short cryptic variable names and you will be viewed as
# - super-intelligent by all non-programmers 
# - super-stupid by all who know how to program
names(academics.cryptic) <- c("pos", "slr", "cit", "wyr")   
# Remember that your scripts are the best (and typically only) documentation
# for your data-analysis. Make them readable and accessible! 
names(academics.cryptic)
```


Creating new variables to data frames 
----------------------------------------------------

Often we need to calculate something based ont the existing variables in the data frame and then include the result in the data frame. 
```{r}
academics$costs <- academics$salaryBeforeTaxes * 1.56

academics$citationsPerYear <- academics$citations / academics$workingYears

academics

```

It does not matter if the variable is already present in the data frame or not. 

```{r}
# Budjet deficits solved!
academics$salaryBeforeTaxes
academics$salaryBeforeTaxes <- academics$salaryBeforeTaxes * 0.80
academics$salaryBeforeTaxes

# Here's how the commands work. On the right side of of the arrow, old values are used 
# to calculate the new values. Then those values are assigned back to the data frame
# using the same variable names. The orginal values are lost.

# costs variable was calculated by multiplying salaryBeforeTaxes by 1.56.
# costs variable is not automatically updated to reflect the new salaries.
academics$new_costs <- academics$salaryBeforeTaxes * 1.56

academics
```

If we made a mistake in creating the new variable, we can "cancel" creating it by assigining NULL to it. This removes the newly created variable from the data frame.

```{r}
academics$citatonsPerCitations <- academics$citations / academics$citations
academics

# The command assigned "emptiness" as the value of the variable, and it's gone!
academics$citationsPerCitations <- NULL

academics
```




There are also other ways to create new variables and incude them in the data frame:

```{r}
academics$citationsPerYear <- NULL

academics <- transform(academics,
          citationsPerYear = citations / workingYears)
academics
```




Creating values conditionally
--------------------------------------

Sometimes it's necessary to create new variables conditionally, conditioned on the values of
other variables

For instance, we could be interested to calculate a new variable which describes whether a person has citations or not:

```{r}
academics$hasCitations <- ifelse(academics$citations > 0, 1 ,0)
academics

```




Comparative operators 
--------------------------------

In the above, we used the comparator operator >. They can be used to compare things, and the results will be either TRUE or FALSE. 

[Operators](https://cran.r-project.org/doc/manuals/r-release/R-lang.html#Operators)

```{r}

5 > 2   # is greater than
5 < 2   # is smaller than
5 == 2  # is equal to
5 != 2  # is not equal to
5 >= 5   # is greater or equal
5 <= 5   # is smaller or equal 

!(5 == 2)  # ! alone is negation

```


Attach and detach
-----------------------

First I will clear the previously created variables so that we do not get confused. 

```{r}
rm(position)
rm(salaryBeforeTaxes)
rm(citations)
rm(workingYears)
```
Try to use those variables! They are not there anymore after removing them. However, the data they contained is still part of the data frame academics which was created. And those data frames do have those variables left, because they are part of the the data frame. 
```{r}
academics$position
```

Notation can be simplified by giving a certain data frame a privileged status using attach:


```{r}
attach(academics)

citationsPerYear <- citations / workingYears

detach(academics)

```

With
-----------------------
Sometimes it can be easier to use the command "with". The idea of "with" is that it defines the data frame from which to look for the variables, but it does not alter the data frame by default. 

```{r}

citationsPerYear <- with(academics.cryptic, cit / wyr)
# citations per year is not added to academics
academics.cryptic

# If you want to add it, write like this:
academics.cryptic$cpy <- with(academics.cryptic, cit / wyr)
academics.cryptic
```


Using some functions
==========================

We can use functions like "mean" and "sum" to calculate things from the vectors.  

```{r}

mean(academics$salaryBeforeTaxes)
sum(academics$citations)
```

However, quite often there are missing values or other invalid data. In these cases, the functions 
quite often return NA = not available or NaN = not a number or something similar. 

```{r}
mean(academics$citationsPerYear)
```

If you want to calculate for example mean ignoring such values, you can use. 
```{r}
mean(academics$citationsPerYear, na.rm=TRUE) # NA? ReMove!
```

Sometimes, it might be useful to use that information for calculation something:

```{r}
academics$valid <- ifelse( is.na(academics$citationsPerYear), 0 ,1)
academics

```



Matrices
===============================

Data frames are 2 x 2 tables, variables always on columns and cases on rows. It is good to know that there is also another way to represent tabular data using matrices. Matrices are R's equivalent to matrices in linear algebra, and they "know" how to become in matrix like fashion with some of the operators. 


```{r}
# First beside one another
# Let's create a couple of vectors

kek <- seq(1,20,2)
kek
kone <- runif(n = 10,min = 1,max=30)
kone

#kone is a vector of random uniformly distributed continuous values, picked from the range [1...30]

kekkone <- cbind(kek,kone)
kekkone

#cbind is a short for "bind columns together", and that's what was done

class(kekkone)
names(kekkone)
```
We have created a matrix. Notice names does not give a vector of variable names. Matrix can usually be transformed into a data frame by saying:

```{r}

kekkone.df <- as.data.frame(kekkone)

kekkone.df
```


If the values in the vectors were the values of ten different values, we could say:

```{r}
konekek <- rbind(kek,kone)
konekek

# rbind is short for "bind rows together", or stack them on top of one another

konekek.df <- as.data.frame(konekek)
konekek.df


names(konekek.df) <- c("Openness", "Conscientciousness", "Extroversion","Agreeableness",
                       "Neuroticism", paste("X",1:5,sep=""))
names(konekek.df)
konekek.df

# Vectors can be combined by c (short for combine)
longvector <- c(kek,kone)


```

Data frame can be easily converted to 2 x 2 matrix. 
```{r}

as.matrix(kekkone.df)

```

A simple statistical test: the t-test
======================================

Let's begin by creating a new data frame of academics who are a bit more well-off than the ones we have so far worked with. 

```{r}
academics.aalto <- academics
academics.aalto$salaryBeforeTaxes <- c(2500,3000,3800,5000,12000,50000)

t.test(academics$salaryBeforeTaxes,academics.aalto$salaryBeforeTaxes,paired = F,var.equal = F)
```

Due to the small sample size, the differences which seem obvious are nonetheless not statistically significant. As you can see, we entered two *vectors* into the t.test function. You can see, by looking at the t-test manual (accessed by typing "?t.test"), the function wants to have (two) vector(s) of data value(s) as input. The function can do different types of t-tests depending on the arguments it receives: for instance, a one-sample t-test would be calculated if only a single vector was given as an input. The other options are similarly described in the documentation of the function. Can you find the descriptions for "paired" and "var.equal"?