---
title: "Reshaping data"
author: "Esko"
date: "10.01.2016"
output: html_document
---


Let's open the satest data again. 
```{r}

satest <- read.table("satest_sample_data.txt", header=TRUE)

```

The video clips can be divided to four groups. The first three videos are practice videos, which should be actually removed from the analysis. After that, the remaining 24 videos are made in sets of 8 videos. 

Half of the participants made learning intervention between the first and second test (test_order == 1), and the other half performed the intervention between the second and third test (test_order == 2). 

The tests were performed in three sets. Now we include that information to the data. 

Recoding
=====================

It is easy to use logical operators to create a new variable set.

```{r}

satest$set <- 3
satest$set[ satest$clip_order < 20] <- 2
satest$set[ satest$clip_order < 11] <- 1
satest$set[ satest$clip_order < 3] <- 0
```

Then we could create an ordered variable or ordered factor.
```{r}
satest$set <- ordered(satest$set)
```

However, there is a good package for recoding which may have a bit more readable syntax. It is up to you what you want to use. 

```{r}
require(car)
 
satest$set <- recode(satest$clip_order, '0:2=0; 3:10=1; 11:19=2; 20:27=3')
satest$set <- ordered(satest$set)
```

We could make the same for the age group variable. 

```{r}

satest$age_grp2 <- recode(satest$age_grp, '"C"="Child"; "A"="Adult"')

```
Notice that age_grp2 is already a factor. Typically R converts strings to factors by default. 

Get rid of the practice clips, set == 0.

```{r}
satest <- subset(satest, set > 0)

# the level 0 is still remembered in the factor definition, drop it as unnecessary
satest$set 

satest$set <- droplevels(satest$set)

```

Calculate the correct answers again.
```{r}

satest$correct <- ((satest$answer == 'notpresent') & (satest$target_type == 'nothing')) |
  ((satest$answer == 'present') & (satest$target_type != 'nothing'))

require(plyr)
corrects <- ddply(satest, .(age_grp2, test_order, set, player_id), summarize,
      correct.mean = mean(correct))

corrects

```

Create a boxplot.
```{r}

boxplot(correct.mean ~ test_order * set * age_grp2, corrects, col=c("red", "blue"))
```



ggplot2 package
===============================

The built-in plotting commands in R are powerful, but may not be very intuitive to use. 
Luckily, there are many packages which provide alternative ways for plotting data. 
ggplot2 is one of my favorites. 

The basic idea is simple: First you map the variable of the data frame to elements of the plot. Then you "add" different representation or "layers" to the plot. 

First, there are example which replicate the previous example using ggplot. However, in order to fully appreaciate its possibility, we need to consider more complex data. 


```{r}

require(ggplot2)

ggplot(corrects, aes(x=set, y=correct.mean, colour=test_order)) + 
  geom_boxplot()


```
But that gives a cryptic error message! I myself run to this accidentally when writing the document, and I had to think a couple of seconds what might be wrong. Then I remembered: ggplot2  -- like many other R functions -- may behave differently depending on weather a variable is factor or not. 

The problem here is that test_order is not a factor and ggplot2 does not know how to use a continuous variable as a basis for color selection in this case.  

```{r}

corrects$test_order <- factor(corrects$test_order)

ggplot(corrects, aes(x=set, y=correct.mean, colour=test_order)) + 
  geom_boxplot()

ggplot(corrects, aes(x=set, y=correct.mean, colour=test_order)) + 
  geom_point()

ggplot(corrects, aes(x=set, y=correct.mean, colour=test_order)) + 
  geom_boxplot() + 
  geom_point()

ggplot(corrects, aes(x=set, y=correct.mean, colour=test_order)) + 
  geom_boxplot() + 
  geom_point()


```

The power of ggplot is that we can always "add"" more thigs to the plot. 

```{r}
ggplot(corrects, aes(x=set, y=correct.mean, colour=test_order)) + 
  geom_boxplot() +
  facet_grid(. ~ age_grp2)

```



```{r}
ggplot(corrects, aes(x=set, y=100*correct.mean, colour=test_order)) + 
  geom_boxplot() +
  facet_grid(. ~ age_grp2) +
  xlab("Set") +
  ylab("Correct (%)") +
  scale_colour_discrete(name="Intervention", labels=c("1-2", "2-3"))

```


We can replace the boxplot with mean and 95 % CI.
```{r}
ggplot(corrects, aes(x=set, y=100*correct.mean, colour=test_order)) + 
  stat_summary(fun.data="mean_cl_normal") +
  facet_grid(. ~ age_grp2) +
  xlab("Set") +
  ylab("Correct (%)") +
  scale_colour_discrete(name="Intervention", labels=c("1-2", "2-3"))

```

Well, that was ugly. Let's try:

```{r}
ggplot(corrects, aes(x=set, y=100*correct.mean, colour=test_order)) + 
  stat_summary(fun.data="mean_cl_normal") +
  facet_grid(test_order ~ age_grp2) +
  xlab("Set") +
  ylab("Correct (%)") +
  scale_colour_discrete(name="Intervention", labels=c("1-2", "2-3"))

```

Or maybe rather:
```{r}
ggplot(corrects, aes(x=set, y=100*correct.mean, colour=test_order)) + 
  stat_summary(fun.data="mean_cl_normal", position=position_dodge(width=0.3)) +
  facet_grid(. ~ age_grp2) +
  xlab("Set") +
  ylab("Correct (%)") +
  scale_colour_discrete(name="Intervention", labels=c("1-2", "2-3"))

```


```{r}
ggplot(corrects, aes(x=set, y=100*correct.mean, colour=test_order)) + 
  stat_summary(fun.data="mean_cl_normal", position=position_dodge(width=0.5), size=1) +
  facet_grid(. ~ age_grp2) +
  xlab("Set") +
  ylab("Correct (%)") +
  scale_colour_discrete(name="Intervention", labels=c("1-2", "2-3")) +
  theme_bw()

```




Reshaping data
==================================

With SPSS, you just quite often forced to reshape data between long and wide format, because the analyzes or plots require the data in either of the format. Based on my experience, R is a bit different. Most of the things which are more intuitive in wide format can be done with long format. Anyway, it is very good to know how to do the trick if needed. 


In the current data, it would be interesting to investigate the correlation between the first and the last set. The data is in long format, and now we want to change it to the wide format. 

It is quite easy to run into problems with reshape. The problems arise from the specification of the correct parameters. I have found it easiest to open the documentation for the reshape and then go through the parameters, including those which seem to apply. 

Long to wide
------------------
```{r}

corrects.wide <- reshape(corrects, 
        v.names="correct.mean", # data
        timevar="set",          # grouping
        direction="wide")
head(corrects.wide)


```

Wide to long
---------------
```{r}
corrects.long <- reshape(corrects.wide, 
        varying=c("correct.mean.1", "correct.mean.2", "correct.mean.3"), # data
        v.names="set",                                                   # grouping
        direction="long")

head(corrects.long)


```

This can be a bit tricky

reshape2 package
-----------------
```{r}


corrects.long2 <- melt(corrects.wide, 
        id.vars = "player_id", 
        measure.vars=c("correct.mean.1.
                       timevar="set",          # grouping
        direction="wide")

# dcast

```


Correlationg between the first and the last set
------------------------------------------------------

In ggplot, the aesthetics must must be mapped on the single row of the data. 


```{r}
ggplot(corrects.wide, aes(x=correct.mean.1, y=correct.mean.3)) + 
  geom_point()

```


It is possible to calculate also the correlation:
```{r}

with(corrects.wide, cor(correct.mean.1, y=correct.mean.3))

with(corrects.wide, cor.test(correct.mean.1, y=correct.mean.3))

```
