---
title: "Testing for association between two continous variables"
author: "Esko, commented by Markus"
date: "22.08.2016"
output: html_document
---


Dataset
-------------------
We will load the dataset created in section _Merging two data frames_, where we combined the dataset containing game performance data with that containing working memory test data. 

```{r}
game.wmc <- read.table("kr_gamewmc.txt", header=T)
```


Question: Is accuracy in the game associated with working memory capacity?
--------------------

Let's calculate _accuracy_ anew:
```{r}
game.wmc$accuracy <- game.wmc$Phit + game.wmc$PcorrectRejection
```

In order to visualize an association between two continous variables, a scatter plot would be perfect. 

```{r}
plot(game.wmc$pcu_score, game.wmc$accuracy) 
```

We can also calculate Pearson correlation (the default for the function _cor_). 

```{r}
cor(game.wmc$pcu_score, game.wmc$accuracy) 
```

But there were two missing values in the working memory test. We can either remove them, or instruct cor function to use only complete observations.

```{r}
cor(game.wmc$pcu_score, game.wmc$accuracy, use="complete.obs") 
```

With cor.test, it is possible to calculate the significance of the correlation, also. Note that missing values are removed by default.

```{r}
cor.test(game.wmc$pcu_score, game.wmc$accuracy)  
```

We could also use a linear regression model to predict accuracy based on the working memory score. 

```{r}
fm <- lm(accuracy ~ pcu_score, game.wmc)
summary(fm)
```

The function _summary_ is useful for inspecting the model. The fitted model can be also given to plot command, which produces plots related to regression diagnostics. 

```{r}
plot(fm)
```

Remember that you can also access the information presented in summary and plot functions programmatically. See the documentation of lm for details. 

```{r}
resid(fm) # residuals using a function resid
fm$resid # residuals using $-notation

```

Finally, we could calculate the F-statistics for all the terms using the _anova_ function. In this case, the relevant F statistic was already available in the summary because the model had only a single predictor. 

```{r}
anova(fm)
```


Association of pcu_score after control for gender
-----------------------------------------------------

We could add categorical variables to the model, also.

```{r}

fm1 <- lm(accuracy ~ pcu_score, game.wmc)
summary(fm1)

fm2 <- lm(accuracy ~ pcu_score + gender, game.wmc)
summary(fm2)
```

From the output, we can see that including the main effect of gender does not make the model better. We can test this explicitly. 

```{r}
anova(fm1, fm2) # by default F-test
anova(fm1, fm2, test="Chisq")
```



