---
title: "merge"
author: "Esko & Markus"
date: "22.08.2016"
output: html_document
---

Merging two data frames 
========================

By going through this example, you will learn how to merge two data frames.

Dataset
----------------

The first dataset is the game data described previously (see _comparing groups_).
```{r}
game <- read.table("kr_game.txt", header=T)
```
The second dataset contains data from a working memory test for the participants. It has the following variables:
- player_id : the participant id
- pcu_score : Percentage Unit Score of a separate counting span working memory test. The higher the score, the higher the working memory capacity

```{r}
wmc <- read.table("kr_wmc.txt", header=T)
head(wmc)
```


Merging by player_id
-------------------------------------------
```{r}
game.wmc <- merge(game, wmc, by="player_id")

head(game.wmc)
```

That was easy. I have named the new data frame "game.wmc"", but any other name like gamewmc, would have been fine, also.

```{r}
write.table(game.wmc, "kr_gamewmc.txt", row.names=F)
```

More complex merges
-------------------------------------------

The above operation was an example of one-to-one merge: Both datasets contain a specific ID on one row only. More complex situations are, however, conceivable: sometimes it's necessary to perform a one-to-many merge or a many-to-many merge. Compared to, say, SPSS, it's super-easy to perform these more complex operations in R. Come talk with us if you need to do this with your data. 