---
title: "Representing data and making simple calculations"
output: html_document
---


Objects are way to represent data in a programming language like R. For different kinds of data, the computer needs different kind of objects. 

Vectors
==========================

In R, the basic object is vector. It is an ordered set of data containing one or more elements. 

When you enter text to the R console, the interpreted will parse your input, and create a vector. The vectors "know" what kind of data they have in them.

[R lang defitions: Vectors](https://cran.r-project.org/doc/manuals/r-release/R-lang.html#Vector-objects)


```{r}

# creates a vector, which contains only one number
3    

# creates a vector with three members
c(7, 6, 3)

# the vectors have type
class(3)
class(c(7, 6, 3))

# and length
length(3)
length(c(7, 6, 3))


# creates object of type character
"Vector"
c("Vector", "says", "hello!")

# beside the type, everything is same as with numbers
class("Vector")
class(c("Vector", "says", "hello!"))

length("Vector")
length(c("Vector", "says", "hello!"))

```

Variables 
=========================

The power of programming language is on the possibility to describe the algorithm (= what will be computed) regardless of the actual data. For this we need variables, which can represent the actual objects in the algorithm. 

```{r}

salaryBeforeTaxes <- 2734.51
employerCosts <- salaryBeforeTaxes * 1.56 
print(employerCosts)

```

Here the R console first creates a vector of type numbers with one member (2734.51). Then it is assigned to the variable salaryBeforeTaxes with operator <-. You can thing a variable a box where you can but data. 


Operators
===========================

Objects represent data and variables can represent them in algorithms. In the previous example, we used multiplication which was represented as *. * is an operator. When an operator encounters objects (or variables which can represent them), new objects are created. 

However, at first there is nothing that exciting if you remember anything about elementary school mathematics. 

[R lang definition: Operators](https://cran.r-project.org/doc/manuals/r-release/R-lang.html#Operators)


```{r}
# Let's create objects and assign them to variables.
a <- 4
b <- 6
c <- -3

a + b
a - b
a * b
a / b

# NB. Order or calculation follows the ordinary mathematical rules
a + b * c    
(a + b) * c  
a + (b * c) 

```

These we are vectors which were of length 1. Everything happens pretty much the same way for vectors which are longer. Everything operator is applied element-wise. 

```{r}
thesisWorker <- 0
researchAssistant <- 1700
doctoralStudent <- 2200
postdoc <- 3200
professor <- 5500
rector <- 25000

salaryBeforeTaxes <- c(thesisWorker, 
                       researchAssistant,
                       doctoralStudent,
                       postdoc,
                       professor,
                       rector)

employerCosts <- salaryBeforeTaxes * 1.56 
employerCosts

```

Variables are a way to describe algorithms irrespective of data. Vectors are a very powerful way to calculate the same thing for multiple values at once. 


Data frames
==============================

Most of the time researchers in behavioural sciences are working with data which are nicely arranged to two dimensional datasets. On the rows, there are observations and on columns there are different variables which describe different aspects of these observations. In R, these are called as data frame objects or just data frames. 


Data frames can be constructed by stacking vectors beside each other or on top of each other


```{r}

position <- c("Thesis worker", 
               "Research assistant",
               "Doctoral student",
               "Postdoc", 
               "Professos", 
               "Rector")
salaryBeforeTaxes <- c(thesisWorker, 
                       researchAssistant,
                       doctoralStudent,
                       postdoc,
                       professor,
                       rector)
workingYears <- c(0, 1, 3, 8, 20, 30)
citations <- c(0, 
               1,
               8,
               30,
               7784,
               2021)
academics <- data.frame(position, salaryBeforeTaxes, citations, workingYears)

```

We can view this nice data frame either printing it to the console or using View. 

```{r}
academics

View(academics)

```



Dots . in the variable or function names
----------------------------------------

In the previous example, we saved the data frame with a variable name which had an ending ".df". It is good to notice that there is nothing magical here. It is just a part of the variable name, and we could have used any other valid variables name. Unlike to many other programming languages like Python, Java or MATLAB, dot (.) is not used for accessing object/struct members. R uses $-symbol for that purpose. 



Accessing the data in data frames
-----------------------------------------

Vectors (or variables) within data frames can be easily accessed and used with $.  

```{r}

academics$salaryBeforeTaxes
```

Using function names, it is also possible to get all the variables names as a vector. 

```{r}
# get the names as a vector
names(academics)  

```
This can be also used to change them. 
```{r}
# set the names from a vector
academics.cryptic <- academics # make a copy of the data frame first

names(academics.cryptic)

# Use short cryptic variable names and you will be viewed as
# - super-intelligent bt all non-programmers 
# - super-stupid by all who know how to program
names(academics.cryptic) <- c("pos", "slr", "cit", "wyr")   
# Remember that your scripts are the best (and typically only) documentation
# for your data-analysis. Make them readable and accessible! 
names(academics.cryptic)
```

Because they were first creates as vectors, they behave like vectors also when they are part of a data frame. 
```{r}
academics$salaryBeforeTaxes * 1.56
```

Creating new variables to data frames 
----------------------------------------------------

Often we need to calculate something from the existing variables in the data frame and then include them to the data frame. 
```{r}
academics$costs <- academics$salaryBeforeTaxes * 1.56

academics$citationsPerYear <- academics$citations / academics$workingYears

academics

```

```{r}
academics$citatonsPerCitations <- academics$citations / academics$citations
academics

# If we made a mistake in creating the new variable, we can "cancel" creating it. 
academics$citationsPerCitations <- NULL

# The command assigned "emptiness" as the value of the variable, and see, it is gone!
academics
```


There are also other ways for that

```{r}
academics$citationsPerYear <- NULL

academics <- transform(academics,
          citationsPerYear = citations / workingYears)
academics
```




Creating values conditionally
--------------------------------------

Sometimes it's necessary to create new variables conditionally, conditioned on the values of
other variables

For instance, we could be interested to calculate a new variables which describes weather a person has citations or not:

```{r}
academics$hasCitations <- ifelse(academics$citations > 0, 1 ,0)
academics

```

Attach and detach
-----------------------
Notation can be simplified by giving a certain data frame a privileged status using attach:

```{r}
academics$citationsPerYear <- NULL # remove it so that it is newly created

attach(academics)

citationsPerYear <- citations / workingYears

detach(academics)

```




Comparative operators 
--------------------------------

In the above, we used comparator operator >. They can be used to compare things, and the results will be either TRUE or FALSE. 

[Operators](https://cran.r-project.org/doc/manuals/r-release/R-lang.html#Operators)

```{r}

5 > 2   # is greater than
5 < 2   # is smaller than
5 == 2  # is equal to
5 != 2  # is not equal to
5 >= 5   # is greater or equal
5 <= 5   # is smaller or equal 

!(5 == 2)  # ! alone is negation

```

Calculating something based on non-missing values: a bit advanced stuff => reorganize
```{r}

#rowMeans(mydata[,paste("q",1:4,sep="")],na.rm=TRUE)

#mydata$mean.missing <- rowMeans(mydata[,paste("q",1:4,sep="")],na.rm=TRUE)
#mydata


```



Matrices
===============================

Data frames are 2 x 2 tables, variables always on columns and cases on rows. It is good to know, that there is also another way to represent tabular data using matrices. Matrices are R's equivalent to matrices in linear algebra, and they "know" how to become in matrix like fashion with some of the operators. 


```{r}
# First beside one another
# Let's create a couple of vectors

kek <- seq(1,20,2)
kek
kone <- runif(n = 10,min = 1,max=30)
kone

#kone is a vector of random uniformly distributed continuous values, picked from the range [1...30]

kekkone <- cbind(kek,kone)
kekkone

#cbind is a short for "bind columns together", and that's what was done

class(kekkone)
names(kekkone)
```
We have created a matrix. Notice names does not give a vector of variable names. Matrix can usually be transformed into a data frame by saying:

```{r}

kekkone.df <- as.data.frame(kekkone)

kekkone.df
```




```

If the values in the vectors were the values of ten different values, we could say:

```{r}
konekek <- rbind(kek,kone)
konekek

# rbind is short for "bind rows together", or stack them on top of one another

konekek.df <- as.data.frame(konekek)
konekek.df


names(konekek.df) <- c("Openness", "Conscientciousness", "Extroversion","Agreeableness",
                       "Neuroticism", paste("X",1:5,sep=""))
names(konekek.df)
konekek.df

# Vectors can be combined by c (short for combine)
longvector <- c(kek,kone)


```

Data frame can be easily converted to 2 x 2 matrix. 
```{r}

as.matrix(kekkone.df)

```


