---
title: "rmANOVA"
author: "Esko"
date: "01.07.2016"
output: html_document
---


Dataset
----------------

We start with the raw game data from the previous sections.

```{r}
game.raw <- read.table("kr_gameraw.txt", header=T)
head(game.raw)
```

The data was about learning game, where each player player 30 video clips. 

Question: Did players performance improve while playing the game? 
-------------------------------------------------------------------

A very straigforward way to examine this question would be to divide the 30 videos clips to three sets, each containing 10 clips. Then we could calculate -- that is, aggregate -- the Phit value by each set and player. 

Recoding video_id to set
--------------------------
There is variable video_id, which coded the order of the videos from 0 to 29. We can use it to form a new variable set. 

```{r}
# Using a quite easy to understand recode function from package car
require(car)
game.raw$set <- recode(game.raw$video_id, "0:9=1; 10:19=2; 20:29=3")

# Making the same thing with some math and using function floor to round downwards. 
# E.g. 7 / 10 = 0.7; floor(0.7) = 0; 0 + 1 = 1
game.raw$set2 <- floor(game.raw$video_id / 10) + 1

# The results are identical. It is up to you which one to use.  
game.raw$set == game.raw$set2
```

Finally, it is a good to convert set explicitly to ordered factor.

```{r}
game.raw$set <- ordered(game.raw$set)
```


Aggregation by player and set
--------------------------------
```{r}
require(plyr)
params <- ddply(game.raw, .(age_grp, player_id, set), summarize,
                  Phit = mean(hit))

head(params)
```

We could also include more parameters to the calculation.

```{r}
params <- ddply(game.raw, .(age_grp, player_id, set), summarize,
                  Phit = mean(hit),
                  PcorrectRejection = mean(correctRejection),
                  accuracy = Phit + PcorrectRejection)
head(params)
```

We calcualte both Phit and PcorrectRejection. Then we calculate accuracy by adding them. This is possible already here, because the variable Phit and PcorrectRejection are created before accuracy. 


In order to answer the question, it would be great to make first a plot and then repeated measures ANOVA. In the next section we will use ggplot2 to make some fancier plots, but let's just stick to the R's in-built plotting tools.  

```{r}
# boxplot
boxplot(accuracy ~ set + age_grp, params)

# and this could be used to summarise the effects using means
interaction.plot(params$set, params$age_grp, params$accuracy)
```

Repeated-measures ANOVA
-------------------------

Making repeated measures ANOVAs in R are is a bit complicated using the in-built functions because you need to understand what you are doing. There I recommed using package ez. 


```{r}
##################################################################################
# Before making anovas with R, set the default contrasts to orthogonal, 
# if want to have results comparable to SPSS. 
#
# Especially, if you want to use type III Sum of Squares like in SPSS by default, 
# you MUST use orthogonal contrasts! Otherwise your results may be wrong. 
##################################################################################
options(contrasts = c("contr.sum", "contr.poly")) 

require(ez)
fm <- ezANOVA(params, 
        dv=accuracy,
        wid=player_id,
        within = .(set),
        between = .(age_grp),
        type=3) # we ask type III Sum of Squares
# If you didn't convert set to a factor, you will run into problems here, because ezANOVA treats it 
# numeric and it confuses it

print(fm)
```

How about the post-hoc testing? Ask the ezANOVA to provide the linear model also, and then feed it to lsmeans. 

```{r}
# I repeat this here: Orthogonal contrasts never forget with type III Sum of Squares! 
options(contrasts = c("contr.sum", "contr.poly")) 

fm <- ezANOVA(params, 
        dv=accuracy,
        wid=player_id,
        within = .(set),
        between = .(age_grp),
        type=3,
        return_aov=T) # get the model

require(lsmeans)

# pairwise post-hocs
lsmeans(fm$aov, pairwise ~ set)
# with summary, you can get e.g. bonferroni or holm adjusted pairwise comparisons
summary(lsmeans(fm$aov, pairwise ~ set), adjust="bonferroni") 
summary(lsmeans(fm$aov, pairwise ~ set), adjust="holm") 


# polynomial contrast
lsmeans(fm$aov, poly ~ set)
```

Because we the main effect of set was significant (and the interaction was not), and the polynomial contrasts indicate positive linear effect, we can conclude that the players are learning.

```{r}
# It is possible to investigate interactions also, but remember that 
# F test for interaction was not significant. So there are for 
# illustration only.

# a bit complicated and makes too many tests
lsmeans(fm$aov, pairwise ~ set*age_grp)

# maybe better
lsmeans(fm$aov, pairwise ~ set | age_grp)
lsmeans(fm$aov, poly ~ set | age_grp)
```