---
title: "select_variables"
author: "Markus"
date: "January 9, 2016"
output: html_document
---

Selecting variables
===============================

```{r results="hide", echo=FALSE, warning=FALSE, message=FALSE}
thesisWorker <- 0
researchAssistant <- 1700
doctoralStudent <- 2200
postdoc <- 3200
professor <- 5500
rector <- 25000

salaryBeforeTaxes <- c(thesisWorker, 
                       researchAssistant,
                       doctoralStudent,
                       postdoc,
                       professor,
                       rector)

salaryBeforeTaxes 

employerCosts <- salaryBeforeTaxes * 1.56 
employerCosts

position <- c("Thesis worker", 
              "Research assistant",
              "Doctoral student",
              "Postdoc", 
              "Professor", 
              "Rector")

workingYears <- c(0, 1, 3, 8, 20, 30)

citations <- c(0, 
               1,
               8,
               30,
               7784,
               2021)
academics <- data.frame(position, salaryBeforeTaxes, citations, workingYears)
```

In the previous exercise, we learned how to choose observations from a data frame. At the same time, we did some variable selection, so we know that selecting observations and selecting variables can be done by indexing the data frame. The subset command could also be used in choosing variables. 

```{r}

academics[1:4]
academics[2:3]

# The names of the variables are saved in a special vector that can be accessed by saying:
names(academics)
```

So the data frame can be indexed by variables using the bracket notation. The brackets can easily become confusing, though, because double brackets, as in "academics[[3]]" have a different meaning than single brackets as in academics[3]. Let's try to make some sense of this:

```{r}
academics[3]
class(academics[3])

#Aha! So the previous command produced a 6*1 slice of the original data frame, which was a data frame #itself. OK...

#What about double brackets?

academics[[3]]
class(academics[[3]])

#OK, so now we got the same numbers, but no longer as a data frame: the double brackets extracted a #numeric vector from the data frame.

#Also:
dim(academics[[3]])
dim(academics[3])
# So, the vector is dimensionless in this sense, and the corresponding slice of the data frame has dimensions 6*1
```

A single variable can be chosen using the following notation:

```{r}
academics["citations"]
```

You can also index the data frame either using positive or negative numbers (negative numbers leave out variables). With negative numbers, the brackets need to be used, like so:

```{r}
academics[-(1:2)]
# which equals
academics[3:length(academics)] # I used the length command here for demonstration purposes
```

```{r error=FALSE}
# Can we then use the colon notation for selecting variables by names?
# academics["salaryBeforeTaxes:workingYears"]
# This doesn't seem to work... So we need to think of something else...

# Let's see if the previously learned subset command might work:
subset(academics,select=salaryBeforeTaxes:workingYears)
# And it does, hooray!
```

That's all folks!