---
title: "Select observations"
author: "Markus"
date: "8 Jul 2016"
output: html_document
---


Selecting observations from a data frame
====================================================

Previously, we created the dataset academics2 with the questionnaire responses
Let's modify the data a bit further

```{r}
# Let's give the academics genders:

academics2$gender <- c("M","M","F","M","F","M")

# The names of the observations are stored in the row names vector that can be accessed using:

row.names(academics2)

# As you can see, the names are enclosed in quotes, so they are of type "character"
# We could also give actual names to the observations, e.g.

row.names(academics2)[1] <- "Cosmo Kiveskoski"
row.names(academics2)[2] <- "Kalja Naulapää"
row.names(academics2)[3] <- "Miina Kenttä"
row.names(academics2)[4] <- "Heppu Kaalinpää"
row.names(academics2)[5] <- "Merikeiju Jumalankiitos"
row.names(academics2)[6] <- "Meirami Pervonkallio"

academics2

# Oops! THere's a mistake in the gender of the rector! Let's correct:

academics2["Meirami Pervonkallio","gender"] <- "F"
academics2

#Names can be changed back to numbers by saying:
row.names(academics2) <- as.character(1:6)

#The result is equivalent to just saying:
row.names(academics2) <- 1:6
row.names(academics2)
# But as we can see, the row names are enclosed in quotes, so they are of the type "character"
# So R implicitly converts the numbers to characters for row names, trying to guess what 
# we want
```

Sometimes it's necessary to create subsets of the data, to select only certain observations.

```{r}
# So how to select, e.g. only males?
# We can do this by indexing the data frame as follows:

academics2[academics2$gender == "M",]

# The latter part actually creates a boolean vector (a vector of TRUE/FALSE values):

academics2$gender == "M"

# It's good to note that R is case-sensitive:

academics2$gender == "m"

# So we are asking the program to select the observations for which the logical expression is TRUE
```

Even though subsetting data frames using indices is a necessary skill to learn while using R,
there is also a perhaps more intuitive command that can be used. It's called *subset* and it's a good option as long as you are not writing functions of your own. It works as follows:

```{r}
subset(academics2,salaryBeforeTaxes > 3000)
```

Here we chose the observations for which the logical test is TRUE. The *subset* command can also be used for selecting only certain variables using the option *select*

```{r}
subset(academics2,salaryBeforeTaxes > 3000, select = 5:8)
```

Then, a couple of exercises:

How could we ask for the summary of all variables for FEMALES?

Are the salaries of the females and males in this representative sample of academics equal?

Back to Indexing
---------------------

Couple of more details of using the indices:

```{r}
academics2[ c(3,4,5,6), ] # is equal to:
academics2[ c(3:6), ] # is equal to:
academics2[ -c(1:2), ]
```

To understand the negative indices, you can think of Venn diagrams and the negation of a set.

We can also use our brand new brilliant row names vector to choose observations. If, for instance, we want to have a look at Miina's values, we can say:

```{r}
academics2[ "Miina Kenttä" , ]

#If you don't like Miina, you can do:

academics2[ row.names(academics2) != "Miina Kenttä" , ]
```

Because the indexing implicitly creates logical vectors, you can also crate one explicitly and use it for choosing the observations:

```{r}
selection <- c(TRUE,TRUE,TRUE,TRUE,FALSE,FALSE)

academics2[selection,]

# The inverse is defined as:

!selection

# so

academics2[!selection,]

# For a quicker notation, ones and zeroes can also be used:

academics2[as.logical(c(1,1,1,1,0,0)),]
```

As a final remark, indexing rows (observations) and columns (variables) can naturally be done simultaneously:

```{r}
academics2[academics2$gender == "F", 5:6]
```
